---
title: "Tetlin Witch Report"
author: Jane Biologist
format: html
fig-align: center
editor: source
---


```{r setup}
#| echo: false
#| message: false

knitr::opts_chunk$set(warning = FALSE, 
                      echo = FALSE,
                      message = FALSE, 
                      fig.retina = 3, 
                      fig.align = "center")
library(unmarked)
library(terra)
library(tidyverse)
library(RColorBrewer)
library(sf)
library(leaflet)
```

```{r load-data}
#| cache: true

# Load site data
dat <- read.csv("../data/csv/dat.csv")

source("../R/simulate_data.R")
source("../R/create_map.R")

# Scaled covariates
sc <- dat |>
    dplyr::select(forest, water) |>
    scale()

# Load site data and scale them
load("../data/rdata/unmarked_df.Rdata")
```

```{r fit-model}
#|cache: true

# Fit single season occupancy model
mod <- fit_model(unmarked_df)
```

## Introduction

Invasive witches have become a management concern at Tetlin National Wildlife Refuge. As such, there is a need to estimate witch occurrence within the Refuge.

## Methods

### Data Collection

We visited a sample randomly distributed sites across Tetlin Refuge. At each site, we spent one hour looking and listening for witches. We revisited each site eight times.

### Model

We estimated witch occupancy and detection using a single-season occupancy model. We used the `unmarked` R package. [blah, blah, blah]


## Results

We surveyed a total of `r nrow(dat)` sites. The average distance to water at our sites was `r round(mean(dat$water), 2)` m. The average distance to forest at our sites was `r round(mean(dat$forest), 2)` m.

```{r}
#| out-width: "50%"
#| fig-cap: "A map of the sites surveyed for witches, Tetlin National Wildlife Refuge, Alaska."

# Import data
tetlin <- sf::st_read("../data/shapefile/tetlin.shp", 
                      quiet = TRUE)
sites <- read.csv("../data/csv/sites.csv")

# Create leaflet map
base_map(sites, tetlin)
```

We observed witches on `r sum(dat[6:13])` of 800 site visits, for a naive occupancy of `r round(sum(dat[6:13])/ncell(dat[6:13]), 2)`. 

```{r plot-psi}
#| fig-height: 3
#| fig-cap: Occupancy of witches at Tetlin National Wildlife Refuge, Alaska, 2024.

# Calculate predicted values (unscaled)
pred <- rbind(plotEffectsData(mod, "state", "forest"), plotEffectsData(mod, "state", "water")) %>%
  mutate(covariateValue = case_when(
    covariate == "forest" ~ covariateValue * attr(sc, 'scaled:scale')[[1]] + attr(sc, 'scaled:center')[[1]],
    covariate == "water" ~ covariateValue * attr(sc, 'scaled:scale')[[2]] + attr(sc, 'scaled:center')[[2]]
  ))

# Plot predicted values (psi)
ggplot(pred, aes(x = covariateValue, y = Predicted), 
  group = covariate) + 
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              linetype = 2, 
              alpha = 0.1) +
  xlab("Distance (m)") +
  ylab("Psi") +
  facet_grid(~covariate, scales = "free")
```